---
title: "NOAA Storm Data Analysis"
author: "Danish Tamboli"
date: "7/21/2020"
output: html_document
---
### Synopsis

#### The NOAA Dataset consists of readings from 1950-2011, with readings being taken more frequently in the later years, Analysis of the NOAA Storm Data is being carried out in hopes of finding the answer to the following questions:

#### Across the United States, which types of events (as indicated in the [EVTYPE] variable) are most harmful with respect to population health?

##### Calculated Taking the following variables into account:
 * EVTYPE
 * FATALITIES
 * INJURIES
 
        
#### Across the United States, which types of events have the greatest economic consequences?

##### Calculated Taking the following variables into account:
 * EVTYPE
 * PROPDMG
 * CROPDMG

## Data Processing
#### Downloading the U.S. National Oceanic and Atmospheric Administration's (NOAA) Storm Dataset.
```{r download}
if(!file.exists("NOAA_Strom_Data.csv.bz2")){
        download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2","NOAA_Strom_Data.csv.bz2")
}
```

#### Reading and Storing the Dataset using read.csv function.
```{r reading,cache=TRUE}
NOAA_Storm_Data <- read.csv("NOAA_Strom_Data.csv.bz2")
```

#### Lets have a look at the first 6 lines of the Dataset and An Overview of the classes of the Collumns present in the dataset.
```{r summary ,cache=TRUE}
head(NOAA_Storm_Data)
str(NOAA_Storm_Data)
```

#### Coverting the $BGN_DATE and $BGN_TIME from Character class to Date and POSIXlt respectively.
```{r datetime,cache=TRUE}
MOD_NOAA_Storm_Data <- NOAA_Storm_Data
MOD_NOAA_Storm_Data$BGN_DATE <- as.Date(NOAA_Storm_Data$BGN_DATE, "%m/%d/%Y")
MOD_NOAA_Storm_Data$BGN_TIME <- strptime(paste(MOD_NOAA_Storm_Data$BGN_DATE,MOD_NOAA_Storm_Data$BGN_TIME),"%Y-%m-%d %H%M")
```

#### We see that now $BGN_DATE and $BGN_TIME are in their correct formats.
```{r datetimee structure}
str(MOD_NOAA_Storm_Data$BGN_DATE)
str(MOD_NOAA_Storm_Data$BGN_TIME)
```

## Results

### Across the United States, which types of events (as indicated in the \color{red}{\verb|EVTYPE|}EVTYPE variable) are most harmful with respect to population health?

#### Finding Total Number of Fatalities and Injuries with respect to Event Type.
```{r total fatalities and injuries,cache=TRUE}
# Checking if INJURIES AND FATALITIES CONTAIN ANY NA Values.
table(is.na(MOD_NOAA_Storm_Data$INJURIES))
table(is.na(MOD_NOAA_Storm_Data$FATALITIES))

# Seeing how many different types of Events Exist
length(table(MOD_NOAA_Storm_Data$EVTYPE))

# Finding The Total number of Injuries and Fatalities with respect to each Event.
total.injuries <- aggregate(INJURIES~EVTYPE,MOD_NOAA_Storm_Data,sum)
total.fatalities <- aggregate(FATALITIES~EVTYPE,MOD_NOAA_Storm_Data,sum)

# Merging Total Injuries and Fatalities into a single Dataframe and sorting it in Descending order, Number of Fatalities given higher importance than Injuries.
total.injuries.fatalities <- merge(total.injuries,total.fatalities,by = "EVTYPE")
total.injuries.fatalities <- total.injuries.fatalities[order(-total.injuries.fatalities$FATALITIES,-total.injuries.fatalities$INJURIES),]

# Picking the Top 5 and changing the colnames to more convinient ones.
total.injuries.fatalities.top5 <- total.injuries.fatalities[1:5,]
colnames(total.injuries.fatalities.top5) <- c("Event.Type","Injuries","Fatalities")

# Table of Total Injuries and Fatalities with respect to Event Type.
knitr::kable(total.injuries.fatalities.top5)
```

```{r total casualties,cache=TRUE}
# Calculating Cumulative Casualties (Fatalities+Injuries),in Descending order with respect to Different Events.
total.harm <- aggregate(FATALITIES+INJURIES~EVTYPE,MOD_NOAA_Storm_Data,sum)
total.harm <- total.harm[order(-total.harm$`FATALITIES + INJURIES`),]

# Selecting the Top 10 Casualties prone areas.
top5.total.harm <- total.harm[1:5,]
colnames(top5.total.harm) <- c("Event.Type","Total.Casualty")

# Table of Total Casualties varying with Different Events
knitr::kable(top5.total.harm)
```

```{r Most Harmful Events Plot}
#Loading the ggplot2 library for plotting
library(ggplot2)

# TOP 5 Most Casualty Causing / Most Harmful Event Types
ggplot(top5.total.harm,aes(Event.Type,Total.Casualty,fill=Event.Type)) + geom_col() + labs(x= "Event Type",y= "Total Casualty",title= "TOP 5 Most Casualty Causing Event Types")
```

#### We see that the Top 5 Most Harmful events are as follow:  
* TORNADO
* EXCESSIVE HEAT
* TSTM WIND
* FLOOD
* LIGHTNING

## Across the United States, which types of events have the greatest economic consequences?

#### Calculating the Property and Crop Damages,along with Total Damages on a Event Type basis.
```{r Economical damages,cache=TRUE}
# Calculating the Total property Damage of different Events.
property.damage <- aggregate(PROPDMG~EVTYPE,MOD_NOAA_Storm_Data,sum)

# Calculating the TOtal Crop Damange of different Events.
crop.damage <- aggregate(CROPDMG~EVTYPE,MOD_NOAA_Storm_Data,sum)

# Combining the Property and Crop Damage into a singe Dataframe.
economical.damage <- merge(property.damage,crop.damage,by="EVTYPE")

# Calculating total Economical Damage.
economical.damage <- cbind(economical.damage,Total.Damage = property.damage[,2] + crop.damage[,2])

colnames(economical.damage)<- c("Event.Type","Property.Damage","Crop.Damage","Total.Damage")

# Sorting the Economical Damage in Descending Order.
economical.damage <- economical.damage[order(-economical.damage$Total.Damage),]

# Filtering out the top 5 monst Economical Damange causing Events.
economical.damage.top.5 <- economical.damage[1:5,]

#Table of Events and the Damages Incured by them
knitr::kable(economical.damage.top.5)
```

```{r Economy Damaging Events Plot}
ggplot(economical.damage.top.5,aes(Event.Type,Total.Damage/1000)) + geom_col(aes(fill=Event.Type)) + labs(x="Type of Event",y="Total Damage Done in 1000's/K USD($)",title = "Top 5 Economical Damage done by Events (Property Damage + Crop Damage)")
```

#### We see that the Top 5 Most Economy Damaging Types of Events are:

* TORNADO
* FLASH FLOOD
* TSTM WIND
* HAIL
* FLOOD
